WSClean: a new implementation of a fast, generic wide-field imager

### 2014-01-18 articles/2014/Article-WSClean/wsclean.tex

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments.

The most commonly used method to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in CASA. This is the fastest generic wide-field imager available, and CASA is the only software package (known to us) that implements the w-projection algorithm. Other packages rely on the faceting algorithm \citep{facetting-cornwell}, which is about an order of magnitude slower and is less accurate \citep{wprojection-cornwell}. The w-projection algorithm has thus been successful so far. However, the available implementations are unable to cope with the data rates from the new generation of wide-field observatories, which are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR). At the MWA, we have seen that it can take up to tens of hours with the w-projection algorithm to image a two-minute MWA observation when the observed field is at low elevation. Moreover, the w-projection implementation in CASA can not handle images of 10 k $\times$ 10 k, which is desired for making full-sky images at the native resolution of the MWA. This will become an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.

We present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation. Currently it performs approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging. The implementation uses the ``w-stacking method'' for correcting the w-terms \citep{widefield-imaging-ska-cornwell} to obtain the increase in speed. Performance can be improved further by combination with another technique: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane. This is somewhat similar to the w-snapshot algorithm proposed in \citet{widefield-imaging-ska-cornwell}. Hence, we can describe the imager as a hybrid between w-stacking and a variation of the w-snapshot algorithm. It also implements the Clean deconvolution procedure (\citealt{hogbom-clean} and various variations). We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.

After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources). GMRT data and -- if available -- LOFAR data will also be used. We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.

@@

An interferometer samples the complex visibility function
\begin{align}\notag
V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \cdot \\ \label{eq:visibility-function}
& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{align}
where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$. We will not discuss calibration, and assume $V$ has been calibrated before imaging. In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ Jones matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem. Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.

@@

We will now mathematically derive the w-stacking technique. First Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{-2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_e - w_0\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel and/or primary beam.

@@

Due to memory constraints it is often necessary to perform multiple passes over the data set. The overhead caused by this is often negligable, but when both the number of passes and the number of samples is large, the algorithm will become dominated by seeking through the data set. Therefore, we add another term $(\eta+1) N_\textrm{samples}$, where $\eta$ defines the involved overhead. This overhead becomes higher when $N_\textrm{wlayers}$, $N_\textrm{imgsize}$ and $N_\textrm{samples}$ are higher or when the available memory is lower, but is $\sim 0$ in common MWA cases.

For each $w$-layer, an inverse FFT operation is performed with a size equal to the image size. If we define $N_\textrm{imgsize}$ the size of the image along one axis, the number of operations involved in this scales with $N_\textrm{wlayers} N^2_\textrm{imgsize} \log N_\textrm{imgsize}$. After each inverse FFT, the direction dependent phase shift needs to be applied, for which a number of operations $\sim$ $N_\textrm{wlayers} N^2_\textrm{imgsize}$ is required, and the time spent on the inverse FFT normally dominates. Finally, the layers need to be added together and will be scaled, which adds a factor relative to $N_\textrm{samples}$ operations, which also is insignificant relative to the inverse FFTs.

This implies that the total number of operations $N_\textrm{op}$ scales in the following way:
\begin{align} \notag
 N_\textrm{op} \sim & N_\textrm{wlayers} \left( N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N^2_\textrm{imgsize} \right) + \\
 & N^2_\textrm{imgsize} + N_\textrm{samples} \left( N^2_\textrm{kernel} + {\eta} + 1 \right).
\end{align}
If we ignore the possible reading overhead, and assume that the number of $w$-layers scales with the image size, a strict bound in Big-O notation is given by
\begin{equation}
\mathcal{O}\left(N^3_\textrm{imgsize} \log N_\textrm{imgsize} + N_\textrm{samples} N^2_\textrm{kernel}\right).
\end{equation}

We can compare this to $w$-projection algorithm \citep{wprojection-cornwell}, for which the total number of operations scales as
\begin{align} \notag
 N_\textrm{op-wproj} \sim & N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N^2_\textrm{imgsize} + \\
 & N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2 + N_\textrm{wplanes},
\end{align}
where $N_\textrm{wkernel}$ is the size of the added kernel to correct the $w$-corrections in $uv$-space as a convolution, instead of as multiplication in $lm$-space. Each sample needs to be convolved with the combined kernel, hence the term $N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2$. There is also time associated with calculating these kernels, and this can be substantial for the MWA. This scales with $N_\textrm{wplanes}$ the number of $w$-projection planes. The constraints on this number are equal to the constraints on $N_\textrm{wlayers}$ in the $w$-stacking algorithm, as described in \S\ref{sec:gridding-w}. The number of pixels in the $w$-kernel is depending on how quickly the kernel approaches zero. This kernel is a function in $u,v$, and hence the larger the angular size of the image, the smaller an individual $u,v$ pixel becomes, and the larger (in pixels) the kernel becomes. Hence, $N_\textrm{wkernel}$ scales with $N_\textrm{imgsize}$. The resulting bound for $w$-projection is:
\begin{equation}
\mathcal{O}\left(N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N_\textrm{samples} N^2_\textrm{imgsize} + N_\textrm{planes} \right).
\end{equation}
By comparing the time complexity between $w$-stacking and $w$-projection, we can conclude that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are dominating. The main parameter that changes this balance is the number of visibilities to be gridded, i.e., the size of the dataset. For larger datasets, $w$-stacking becomes more favourable. In section TODO we will determine which method is faster in practice for different parameters.

@@

\section{The WSClean imager implementation}

@@

>>>>

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments \citep{perley-noncoplanar-arrays}.

There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking\footnote{TODO when/where was this first mentioned in literature?} \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}. The w-snapshot algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.

Probably the most commonly used software implementation to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in the Common Astronomy Software Applications (CASA, \citealt{casa}), because of its many features, high quality and free availability. New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR). At the MWA we have seen that, when the observed field is at low elevation, it can take up to tens of hours with CASA's w-projection algorithm to image a two-minute MWA observation. The problem of wide-field imaging will become an even bigger challenge when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required. In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-stacking algorithm is  the best approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation. Currently it performs approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging. The implementation uses the ``w-stacking method'' for correcting the w-terms \citep{widefield-imaging-ska-cornwell} to obtain the increase in speed. Performance can be improved further by combination with another techniques: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane. This is somewhat similar to the w-snapshot algorithm proposed in \citet{widefield-imaging-ska-cornwell}. Hence, we can describe the imager as a hybrid between w-stacking and a variation of the w-snapshot algorithm. It also implements the Clean deconvolution procedure (\citealt{hogbom-clean} and various variations). We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.

TODO. After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources). GMRT data and -- if available -- LOFAR data will also be used. We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.

@@

An interferometer samples the complex visibility function
\begin{align}\notag
V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \cdot \\ \label{eq:visibility-function}
& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{align}
where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$. We will not discuss calibration, but assume $V$ has been calibrated before imaging. In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ Jones matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem. Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.

@@

To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{-2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_e - w_0\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel and/or primary beam. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_0$ to the minimum absolute $w$-value, which requires at least two times less layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence is always performed as a complex-to-complex transform.

@@

Due to memory constraints it is often necessary to perform multiple passes over the data set. The overhead caused by this is often negligable, but when both the number of passes and the number of samples is large, the algorithm will become dominated by seeking through the data set. Therefore, we add another term $(\eta+1) N_\textrm{samples}$, where $\eta$ defines the involved overhead. This overhead becomes higher when $N_\textrm{wlayers}$, $N_\textrm{pixels}$ and $N_\textrm{samples}$ are higher or when the available memory is lower, but is $\sim 0$ in common MWA cases.

For each $w$-layer, an inverse FFT operation is performed with a size equal to the image size. If we define $N_\textrm{pixels}$ the number of pixels in the image along each dimension, the number of operations involved in the inverse FFTs scales with $N_\textrm{wlayers} N^2_\textrm{pixels} \log N_\textrm{pixels}$. After each inverse FFT, the direction dependent phase shift needs to be applied, for which a number of operations $\sim$ $N_\textrm{wlayers} N^2_\textrm{pixels}$ is required, and the time spent on the inverse FFT normally dominates. Finally, the layers need to be added together and will be scaled, which adds a factor relative to $N_\textrm{pixels}$ operations, which also is insignificant relative to the inverse FFTs.

This implies that the total number of operations $N_\textrm{op}$ scales in the following way:
\begin{align} \notag
 N_\textrm{op} \sim & N_\textrm{wlayers} \left( N^2_\textrm{pixels} \log N_\textrm{pixels} + N^2_\textrm{pixels} \right) + \\
 & N^2_\textrm{pixels} + N_\textrm{samples} \left( N^2_\textrm{kernel} + {\eta} + 1 \right).
\end{align}
If we ignore the possible reading overhead, and assume that the number of $w$-layers scales with the image size, a strict bound in Big-O notation is given by
\begin{equation}
\mathcal{O}\left(N^3_\textrm{pixels} \log N_\textrm{pixels} + N_\textrm{samples} N^2_\textrm{kernel}\right).
\end{equation}

We can compare this to $w$-projection algorithm \citep{wprojection-cornwell}, for which the total number of operations scales as
\begin{align} \notag
 N_\textrm{op-wproj} \sim & N^2_\textrm{pixels} \log N_\textrm{pixels} + N^2_\textrm{pixels} + \\
 & N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2 + N_\textrm{wplanes},
\end{align}
where $N_\textrm{wkernel}$ is the size of the added kernel to correct the $w$-corrections in $uv$-space as a convolution, instead of as multiplication in $lm$-space. Each sample needs to be convolved with the combined kernel, hence the term $N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2$. There is also time associated with calculating these kernels, and this can be substantial for the MWA. This scales with $N_\textrm{wplanes}$, the number of $w$-projection planes. The constraints on this number are equal to the constraints on $N_\textrm{wlayers}$ in the $w$-stacking algorithm, as described in \S\ref{sec:gridding-w}. The number of pixels in the $w$-kernel is depending on how quickly the kernel approaches zero. This kernel is a function in $u,v$, and hence the larger the angular size of the image, the smaller an individual $u,v$ pixel becomes, and the larger (in pixels) the kernel becomes. Hence, $N_\textrm{wkernel}$ scales with $N_\textrm{pixels}$. The resulting bound for $w$-projection is:
\begin{equation}
\mathcal{O}\left(N^2_\textrm{pixels} \log N_\textrm{pixels} + N_\textrm{samples} N^2_\textrm{pixels} + N_\textrm{planes} \right).
\end{equation}
By comparing the time complexity between $w$-stacking and $w$-projection, we can conclude that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are dominating. The main parameter that changes this balance is the number of visibilities to be gridded, i.e., the size of the dataset. For larger datasets, $w$-stacking becomes more favourable. In section TODO we will determine which method is faster in practice for different parameters.

@@

TODO: discuss low w-values means less-direction-invariant PSF.

\section{The WSClean imager implementation}
We have implemented the $w$-stacking algorithm in the C++ language. Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning. In addition, intrinsics were used for the latter. This speeds up a minor Cleaning iteration,

@@

\section{Discussion}
TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)
